---
title: "R Notes for Beginners"
author: "Marie Rivers"
date: "8/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
# continue with file: fish-length-weight in day 1 folder and fix inserted Lyra pic
* if this file gets too long to be useful, create docs for each header (or find a way to expand and condenes sections)

# Setup and Getting Starting
In an Rmarkdown file, add packages to the setup chunk below knitr. These packages may include: 
`library(tidyverse)`

# Creating Vectors, Matrices, and Data Frames
### data frames
```{r}
# create a data frame
df_name <- data.frame("variable_1", "variable_2")

```
### sequences
```{r}
# create a vector from x to y by z
seq_1 <- seq(from = 1, to = 20, by = 1)
seq_1 # return the vector
```


# Tidy Data & Wrangling

# Filtering and Selecting from Data Frames

# gg plot

- basics
- formating
- cool/fun

- tell it you're using ggplot
- tell it what data you are using and what data you are plotting 
- map a variable to an aesthetic
- tell the graph the type of geometry to create
```{r}

```


# Functions and Loops

# Miscellaneous

# Symbols & Math Operators
### Logarithms `log()`
```{r}
log(4)

```
### exponents `exp()`
```{r}
exp(5)
log(exp(1))
```

# Shortcut Keys
Add a new code chunk: command + option + i

pipe %>%:  command + shift + m


# Rmarkdown formatting

Use back quotes (shift + the button in the top left of the keyboard below esc) to make text appear in code font

# `# Largest header`

## `## medium header`

### `### small header`

#### `#### even smaller header`

##### `##### smallest header`

regular text

*`*italicized font*`*

**`**bold font**`**

- `- bullet points`
  - `"tab" - for indented bullet`
    - `"tab" "tab' - for more indented`

# FIX THIS
Here's a picture of Lyra
 
 ! [Alt text] (/Users/marierivers/Documents/UCSB Environmental Data Science/r_notes/figs/IMG_1280.jpq)

# Debugging and Troubleshooting

# To Do List
- notes from notebook
- Rmarkdown files from EDS 212
- Rmarkdown files from EDS 221
- EDS 212 class lecture slides
  - day 1
  - day 2
  - day 3
  - day 4
  - day 5
- EDS 221 class lecture slides
  - day 1
  - day 2
  - day 3
  - day 4
  - day 5
  - day 6
  - day 7
  - day 8
  - day 9
  - day 10
- EDS 212 interactive sessions
- EDS 221 interactive session
