---
title: "R Notes for Beginners"
author: "Marie Rivers"
date: "8/14/2021"
output:  
  html_document:  
    theme: paper  
    toc: true  
    toc_float: true  
    number_sections: true  
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(here)
library(skimr)
library(GGally)
library(lubridate)
library(janitor)
library(kableExtra)
```
# continue with file: 
* if this file gets too long to be useful, create docs for each header (or find a way to expand and condenessections)

# Setup and Getting Starting
In an Rmarkdown file, add packages to the setup chunk below knitr. These packages may include: 
`library(tidyverse)`
`library(palmerpenguins)`

# Creating Vectors, Matrices, and Data Frames
### data frames
```{r}
# create a data frame
df_name <- data.frame("variable_1", "variable_2")

df_1 <- data.frame(region = c("A", "B", "A", "D"),
                   species = c("otter", "great white", "sea lion", "gray whale"),
                   count = c(12, 2, 36, 6))
df_1
class(df_1$species)
max_count <- max(df_1$count) # get the maximum value of the specified column
```
### sequences
```{r}
# create a vector from x to y by z
seq_1 <- seq(from = 1, to = 20, by = 1)
seq_1 # return the vector
```
### vectors
```{r}
# Create a vector!
# the `c` function is for combind or concatenate
vec_a <- c(4, 5) #this is how you create a vector in R
vec_a #return vector a
vec_b <- c(-3, 10)
# vector additon
vec_c <- vec_a + vec_b
vec_c

# vector subtraction
vec_d <- vec_b - vec_a
vec_d

# vector scalar multiplier:
vec_e <- 4 * vec_b
vec_e

# if we use the asterisk:
vec_a * vec_b

# Dot product %*%
vec_a %*% vec_b
# the dot product should be a single number
```
## Checking data types
```{r}
vec_1 <- c(2, 5, 9, 10, 8, 12, 1, 0)
class(vec_1) # determine the class of a vector
typeof(vec_1) # determine the type of numeric
vec_1_e3 <- vec_1[3] # store the value of the item in the 3rd position of the vector
vec_1_e5to7 <- vec_1[5:7] # store the items in positions 5 through 7
vec_1_char <- as.character(vec_1) # creates a vector where the items are converted from numbers to characters
vec_2 <- c(town = "Santa Barbara", location = "Rincon", swell = "south") # create a vector with variable names and values
```

### Matrix
```{r}
cats <- c(1, 3, 5, 8, 10, 0.2, 4, 8, 10, 11, 100)

# wrap cats into a matrix
cat_matrix <- matrix(data = cats, nrow = 4, ncol = 5, byrow = TRUE)

dogs <- matrix(data = c(1, 0, 2, 3), nrow = 2, ncol = 2, byrow = TRUE)
fish <- matrix(data = c(0, 8, 2, 6), nrow = 2, ncol = 2, byrow = TRUE)
dogs + fish

fish - dogs
4 * dogs
dogs %*% fish

## Leslie Matrix
insect_leslie <- matrix(data = c(0, 0, 600, 0.2, 0, 0, 0, 0.08, 0.6), nrow = 3, ncol = 3, byrow = TRUE)

# Specify the initial conditions
# What is the popluation in each lifestage at year 0?
insect_y0 <- c(12000, 700, 500)

# project the populaiton in each lifestage for year 1. Use the dot project operator to do matix multiplication with the vector of initial conditions
insect_y1 <- insect_leslie %*% insect_y0
insect_y1

#Project the population at each lifestage for years 2, 3, and 4
insect_y2 <- insect_leslie %*% insect_y1
insect_y2

insect_y3 <- insect_leslie %*% insect_y2
insect_y3

insect_y4 <- insect_leslie %*% insect_y3
insect_y4
# if 60% of adult females live after laying eggs, change the last number in the list to 0.6
```
### Lists in R
```{r}
# vectors must contain all the same types (homogeneous). Lists can have different classes (heterogeneous)
urchins <- list("blue", 2, FALSE, 1200L)

# the returned [[1]] refers the 1st element in the list, [[2]] refers to the 2nd element
urchins[[1]] # this returns the contents of that element. Use the double brackets most of the time
urchins[1] # this returns a list that only contains 1 element and that elements is "blue"

urchins_2 <- list("blue", 
                  c(1, 2, 3), 
                  seq(from = 0, to = 50, by = 0.5),
                  1200L)

tacos <- list(topping = c("onion", "cilantro", "guacamole"),
              filling = c("beans", "meat", "veggies"),
              price = c(6.75, 8.25, 9.50))

taco_price <- tacos[[3]]
tacos$filling
```



# Tidy Data & Wrangling
- reshape data with `pivot_longer()`  
- separating or uniting columns

Use `tidyr::pivot_loger()` to gather columns (convert from wide to long format)  
Use `tidyr::pivot_wider()` to spread columns (convert from long to wide format)
Use `tidyr::separate()` to separate elements in a single column into multiple columns, using a separator you specify, so that each variable is in a single column.  
Everything in the same column should have the same units

## Read in a data csv
```{r}
toolik_chlorophyll <- read_csv(here("data", "toolik_chlorophyll.csv"), na = ".")
```


## Basic data summaries
```{r}
names(diamonds) # returns column names of a data frame
dim(diamonds) # returns the number of rows and columns in a data frame
summary(diamonds) # returns min, 1st quartile, median, mean, 3rd quartile, and max for each numeric column; returns counts for text columns
head(diamonds) # returns the first few rows of the data frame
tail(diamonds) # returns the last few rows of the data frame
skimr::skim(diamonds) # returns a summary of the data
GGally::ggpairs(diamonds) # graphs every numeric variable against every other numeric variable
View(diamonds) #let's you view a data frame
class(penguins$species) # deterimine the data type of a given column

toolik_chlorophyll <- clean_names(toolik_chlorophyll) # makes column names lower case and removes spaces between words
```
## Find the mean, variance and standard deviation of character height
```{r}
sw_height_mean <- mean(starwars$height)
# the default in the mean function is to NOT remove NA values

sw_height_mean_removeNA <- mean(starwars$height, na.rm = TRUE)
# use ?starwars in the console to more info about the dataset and the units of height

# variance of character height (var())
sw_height_variance <- var(starwars$height, na.rm = TRUE)

# standard deviation of character height (sd())
sw_height_sd <- sd(starwars$height, na.rm = TRUE)
```
#### add text that includes a calculated output  
Mean character height is `r round(sw_height_mean_removeNA, 1)` cm with a standard deviation of `r round(sw_height_sd, 1)` cm.

## janitor::get_dupes()
will check for duplicates across all variables unless you specify variables
```{r}
starwars_dupes <- get_dupes(starwars)
starwars_dupes_names <- starwars %>% 
  get_dupes(name)

starwars_dupes_height <- starwars %>% 
  get_dupes(height)
```
### dplyr::across()
allows you to perform operations across multiple columns based on conditions you specify about the columns
```{r}
# convert anything that's a character to lower case
sw_across <- starwars %>% 
  mutate(across(where(is.character), tolower))
```
Find the mean of all numeric columns
```{r}
starwars %>% 
  group_by(homeworld) %>% 
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```
Separate a column into multiple columns based on a delimiter
```{r}
starwars %>% 
  separate(skin_color, into = c("col_1", "col_2"), sep = ",")
```

```{r}
penguins_column_changes <- penguins %>% 
  mutate(across(where(is.numeric), as.character)) %>% # change all numeric variables to characters
  mutate(island = fct_relevel(island, c("Dream", "Biscoe", "Torgersen"))) %>% # order items into factor levels
  unite(island_yr, c(island, year), sep = "_") %>% # combine columns
  filter(species %in% c("Adelie", "Chinstrap")) %>% # filter items included in a vector (but not necessarily in that order)
  relocate(sex, .after = species) %>% # relocate columns
  rename(penguin_spp = species) # rename a column header
```

Storing specific things from a data frame
```{r}
mack_verts <- read_csv(here("data", "AS00601.csv")) %>% 
  clean_names()

### store the 5th value in column "weight" as `mc_wt_5`
mc_wt_5 <- mack_verts$weight[5]
# store the values in position 8 through 20
mc_length_8_20 <- mack_verts$length1[8:20] #
### store everything in column `sampledate` as a vector called mc_dates
mc_dates <- mack_verts$sampledate
```
## Make a salamander subset
```{r}
mc_salamanders <- mack_verts %>% 
  filter(species == "DITE")

mc_all_salamanders <- mack_verts %>% 
  filter(species != "ONCL")

mc_rhol <- mack_verts %>% 
  filter(species == "RHOL")

```


# Filtering and Selecting from Data Frames
```{r}
# use filter to create subset of data based on certain rows. Use pipe operator to perform things in sequence (and then)
# pipe shortcut: command + shift + M
adelies <- penguins %>% # start with penguins then do this... Use two equal signs because you are asking "is equal?"
  filter(species == "Adelie") %>%  # don't need to write the data frame because you already piped it in
  mutate(body_mass_kg = body_mass_g / 1000) # used one equal sign because we are assigning, not asking if equal
unique(adelies$species) # check that the subset only has Adelies
```
## summary statistics for a filtered data frame
```{r}
toolik_biochem <- read_csv(here("data", "2011_Kling_Akchem.csv"), na = ".") %>% 
  janitor::clean_names()

inlet_biochem <- toolik_biochem %>% 
  filter(site == "Toolik Inlet" | site == "Toolik inlet") %>% # filter using OR
  select(p_h, doc_u_m, tdn_u_m) # select specific columns

inlet_biochem_summary <- inlet_biochem %>% 
  summarize(mean_ph = mean(p_h, na.rm = TRUE), # don't include blank values
            mean_doc = mean(doc_u_m, na.rm = TRUE),
            mean_tdn = mean(tdn_u_m, na.rm = TRUE))

mean_biochem <- vector(mode = "numeric", length = ncol(inlet_biochem))
for (i in 1:ncol(inlet_biochem)) {
  mean_val <- mean(inlet_biochem[[i]], na.rm = TRUE)
  mean_biochem[[i]] <- mean_val
}
```



# Conditional statements & for loops
## Conditional statements
```{r}
pm2_5 <- 248
if (pm2_5 < 100) {
  print("low to moderate risk")
} else if (pm2_5 >= 100 & pm2_5 < 150) {
  print("unhealthy for sensitive groups")
} else {
  print("health risk present")
}

species <- "blue whale"
if (str_detect(species, "whale")) {
  print("You found a whale!")
} else {
  print("better luck next time")
}

max_airtemp_c <- 24.1
if (max_airtemp_c > 27) {
  print("temperature too high")
} else {
  print("temperature OK")
}

base_burrito <- 6.50
main_ingredient <- "veggie"
switch(main_ingredient,
       "veggie" = print(base_burrito),
       "chicken" = print(base_burrito + 3.00),
       "steak" = print(base_burrito + 3.25))

# type ?str_detec in the console to learn more. It's in the stringr package which is in the tidyverse
sbc_observations <- c("blue whale", "gray whale", "whale", "gray wale", "dolphin", "sea lion", "blue whale")
str_detect(sbc_observations, pattern = "whale")

# an example of str_detect() within a conditional statement
phrase <- "I reallly love data science!"
if (str_detect(phrase, "data")) {
  print("cool")
}

# if-else statements
pika <- 45
if (pika > 60) {
  print("large pike")
} else {
  print("small pika")
}
```
## For loops
```{r}
fish <- c(8, 10, 12, 23)
total_fish <- sum(fish)
percent_pop <- vector(mode = "numeric", length = length(fish))
for (i in seq_along(fish)) {
  fishy_fish <- 100 * fish[i] / total_fish
  percent_pop[i] <- fishy_fish
}
percent_pop

month.name
months <- c(month.name)
seq_along(months)

for (i in seq_along(months)) {
  print(paste(months[i], "is month", i))
}
```

```{r}
species <- "dog"
switch(species, 
       "cat" = print("meow"),
       "lion" = print("roar"),
       "owl" = print("screech"),
       "fox" = print("obnmoxious"),
       print("something else"))

vec <- seq(from = 0, to = 12, by = 0.5)
length(vec)

bats <- c(0, -1, 2, 1000, 500, 6)
seq_along(bats)

tigers <- c(29, 34, 82)
lions <- c(2, 18, 6)
big_cats <- vector(mode = "numeric", length = length(tigers))

for (i in seq_along(tigers)) {
  total_cats <- tigers[i] + lions[i]
  big_cats[i] <- total_cats
}
big_cats
seq_along(tigers)

animal <- c("cat", "dog", "dog", "zebra", "dog")
for (i in seq_along(animal)) {
  if (animal[i] == "dog") {
    print("I love dogs")
  } else
    print("these are other animals")
}

mean_mtcars <- vector(mode = "numeric", length = ncol(mtcars))

for (i in 1:ncol(mtcars)) {
  mean_val <- mean(mtcars[[i]], na.rm = TRUE)
  mean_mtcars[[i]] <- mean_val
}
mean_mtcars
# ncol() returns the number of columns in a data frame
```

```{r}
animal <- c("cat", "dog", "dog", "car", "zebra", "elephant")
for (i in seq_along(animal)) {
  if (animal[i] == "dog") {
    print("I love dogs")
} else {
    print("this is a different animal")
  }
}

species <- c("dog", "elephant", "goat", "dog", "dog", "elephant")
age_human_years <- c(3, 8, 4, 6, 12, 18)
animal_ages <- vector(mode = "numeric", length = length(species))

for (i in seq_along(species)) {
  if (species[i] == "dog") {
    animal_ages <- age_human_years[i]*7
  } else if (species[i] == "elephant") {
    animal_ages <- age_human_years[i] * 0.88 
  } else if (species[i] == "goat") {
      animal_ages <- age_human_years[i] * 4.7
  }
  animal_ages[i] <- animal_ages
}
```
## Functions and Loops
```{r}
weight_grams <- function(a, b, length_cm) {
  weight_grams = (a * (length_cm^b))
}
barracuda <- weight_grams(a = 0.0181, b = 3.27, length_cm = 60)
print(barracuda)

fish_length <- seq(from = 0, to = 200, by = 1)

barracuda_weight <- weight_grams(a = 0.0181, b = 3.27, length_cm = fish_length)
barracuda_df <- data.frame(fish_length, barracuda_weight)

ggplot(data = barracuda_df, aes(x = fish_length, y = barracuda_weight)) +
  geom_point(color = "red") +
  labs(x = "fish length (cm)",
       y = "weight (grams)",
       title = "Barracuda Weight",
       subtitle = "barracuda weight as a function of fish length")
```

# gg plot
- tell it you're using ggplot
- tell it what data you are using and what data you are plotting 
- map a variable to an aesthetic
- tell the graph the type of geometry to create
```{r}
ggplot(data = diamonds, aes(x = price, y = carat)) +
  geom_point() 
```

```{r}
ggplot(data = diamonds, aes(x = carat)) +
  geom_histogram() # histograms only need 1 variable
ggplot(data = diamonds, aes(y = carat)) +
  geom_histogram() # difine the y-axis to have horizaontal bars

# make a histogram of flipper lenghts
ggplot(data = penguins, aes(x = flipper_length_mm)) +
  geom_histogram() +
  facet_wrap(~species)

ggplot(data = adelies, aes(x = flipper_length_mm, y = body_mass_kg)) +
  geom_point(color = "blue") +
  geom_line(aes(color = island)) + # connect points with lines
  facet_wrap(~island)
```

```{r}
eq_fx <- function(x) {
  (2.4 * (exp(-0.05*x)) * sin(x))
}
ggplot(data = data.frame (x = c(-.5, .5)), aes(x=x)) +
  stat_function(fun = eq_fx)
```

```{r}
# Penguin bill length vs. bill depth
ggplot(data = penguins, aes(x = bill_depth_mm, y = bill_length_mm)) +
  geom_point(aes(color = species)) +
  labs (x = "bill length (mm)",
        y = "bill depth (mm)",
        title = "Palmer penguins bill size",
        caption = "data collected by KB Gorman ey al. at Palmer Station",
        color = "species")
```
Jitter Plot
```{r}
ggplot(data = penguins, aes(x = species, y = flipper_length_mm)) +
  geom_jitter() +
  labs (x = "species",
        y = "flipper length (mm)",
        title = "penguin flipper length by species")
```
change point size based on a variable, make points transparent, reverse y-axis
```{r}
chlorophyll_plot <- ggplot(data = toolik_chlorophyll, aes(x = active_chl_a_ug_l, y = depth)) +
  geom_point(aes(size = active_chl_a_ug_l),
             color = "cyan",
             alpha = 0.5) +  # makes things transparent
  theme_dark() +
  scale_y_reverse()
```
`ggsave` to saved graphs as figures
```{r}
ggsave(here("figs", "chlorophyll_graph.jpg"), plot = chlorophyll_plot)
```

```{r}
toolik_weather <- read_csv(here::here("data", "toolik_weather.csv"), 
                           na = "#N/A") %>% 
  clean_names()
toolik_temp_plot <- ggplot(data = toolik_weather, aes(x = as.factor(month), y = daily_air_temp_mean_c)) + # months in x-axis as factor
  geom_jitter(aes(color = daily_air_temp_mean_c), show.legend = FALSE) + # don't show legend
  scale_color_gradient(low = "blue", high = "orange") +
  theme_minimal() +
  labs(title = "Toolik Station daily air temperature",
       x = "Month",
       y = "Daily mean air temperatures (Celsius)")

toolik_temp_plot
ggsave(here("figs", "toolik_temp_plot.png"), height = 6, width = 7) # set dimensions of exported figure
```



# Symbols & Math Operators

## Logical Expressions
```{r}
5 < 10 # true
# greater than and less then work alphabetically (a is small, z is large)
"cat" > "armadillo" # true
"cat" > "elephant" # false

10 <= 11 # true
5 >= 4 # true
3 <= 1 # false

# play around with vectors
tarantulas <- c(1, 3, 7, 10, 15)
tarantulas > 5 # false, false, true, true, true

pinyon_pine <- 14
pinyon_pine == 10 # does pinyon_pine match a value of 10?
pinyon_pine != 25 # does pinyon_pine NOT match a value of 25
```


## Logical operators:

- `!`
- `&`
- `|`
```{r}
3 < 1 | 8 > 2

6 < 10 & 5 < 1

7 != 11

4 != 4 | 2 < 5 | 6 > 10

0 == TRUE
0 == FALSE
1 == TRUE #true
# zero is false and 1 is true
```

### Logarithms `log()`
```{r}
log(4)
```
### Exponents `exp()`
```{r}
exp(5)
log(exp(1))
```
### Derivatives
```{r}
fx <- expression(x^2)
df_dx <- D(fx, name = 'x')

# Evaluate the slope of f(x) at x = 831
# x <- 831
# evaluate over a range
x <- seq(from = -10, to = 10, by = 0.5)
eval(expr = df_dx)
x <- 831
eval(expr = df_dx)
```

```{r}
ft <- expression(3 * (t ^ 2) - 4 * cos(t))
df_dt <- D(expr = ft, name = 't')
df_dt
```

```{r}
# First Order univariate derivatives
fx <- expression(x ^ 2)
df_dx <- D(expr = fx, 'x')
df_dx
x <- 3
eval(expr = df_dx)

Pa <- expression(2 * (((3 * a) + 2) ^ 4) - 5)
dP_da <- D(expr = Pa, 'a')
dP_da
a <- 1.2
eval(expr = dP_da)

gt <- expression((-4 * t) / (((t ^ 2) + 1)  ^ 3))
dg_dt <- D(expr = gt, 't')
dg_dt
t <- 0
eval(expr = dg_dt)
```

```{r}
# Higher order univariate derivatives
Gz <- expression(4.6 - 1.9 * (z^3))
dG_dz <- D(expr = Gz, 'z')
dG_dz

d2G_dz2 <- D(expr = dG_dz, 'z')
d2G_dz2
```

```{r}
# partial derivatives
f_xyz <- expression(2*x^2*y - 3*x + 1 + 4*z^4*x)
df_dx <- D(expr = f_xyz, 'x')
df_dx

df_dy <- D(expr = f_xyz, 'y')
df_dy

dg_dz <- D(expr = f_xyz, 'z')
dg_dz
```

# Shortcut Keys
Add a new code chunk: command + option + i

pipe %>%:  command + shift + m

run code: command + return
highlight specific lines of code to only run those (command + return), but don't highlight pipes or plus signs


# Rmarkdown formatting  
Use back quotes (shift + the button in the top left of the keyboard below esc) to make text appear in code font

# `# Largest header`  
## `## medium header`  
### `### small header`  
#### `#### even smaller header`  
##### `##### smallest header`  
regular text  
*`*italicized font*`*  
**`**bold font**`**  
- `- bullet points`  
  - `"tab -" for indented bullet`  
    - `"tab tab -" for more indented`

End a line with two spaces to start a new line.`  `  
Or end with a backslash to make a new line `\`\

```{r, echo = FALSE}
# this overrides the global echo option established up top. 
# if echo = false you wont see raw code in the knitted html.
```

Formatted equations: `equation: $e^2 + 1 = 0$`  
equation: $e^2 + 1 = 0$
Prey equation: 
$$\frac{dx}{dt}=\alpha x-\beta xy$$
### Update theme in setup chunk  
---  
title: "My Fancy Rmd"  
subtitle: "Some changes in YAML"  
author: "Marie Rivers"  
output:  
  html_document:  
    theme: paper  
    toc: true  
    toc_float: true  
    number_sections: true  
    code_folding: hide  
---  
add layers to the output line. YAML is space sensitive. put cursor to html, hit return, hit tab...OR, add colun : after html_document and then put cursor infront of html and hit return to automatically get the tab. Google to get different built in Rmarkdown themes

## Code chunk options
```{r}
# adding include = FALSE hides the code chunk when you knit
# ECHO = FALSE hides code
# to hide warnings in knitted output, set warning = FALSE in the beginning of the code chunk
# eval = FALSE means the code wont run (might not let you knit)
```

# Debugging and Troubleshooting

# Miscellaneous
To use git hub, in the console, type usethis::use_git(), then answer yes, they type usethis::use_github(). There might be a more efficient way to add git with a library or something like that.

To upload to github, knit, select the two files, commit, add a comment, commit, pull, close, push

# To Do List
- notes from notebook
- Rmarkdown files from EDS 212
- Rmarkdown files from EDS 221
- EDS 212 class lecture slides
- EDS 221 class lecture slides
  - day 4
  - day 5
  - day 6
  - day 7
  - day 8
  - day 9
  - day 10
  - day 6 interactive
  - favorite dessert
  - icecreamsealion
  - long term ice
  - marie alex meds
  - stl lead
- EDS 221 interactive session
- look at Rstudio cheat sheets and reference guides
- tidyr cheat sheet
